Symbolic Weighted Language Models, Quantitative Parsing and Verification\\ over Infinite Alphabets

### 2021-10-15 algo-best-lemma.tex
intro & abstract
submission FoSSaCS

% !TEX root = main.tex
%

@@

\noindent
If $\pi = v_0,\ldots, v_n$ with $n \geq 1$, 
let us do a case analysis on the edge $v_{n-1} \to v_n$.
%
We suppose Lemma ~\ref{algo-correct} is true for a safe path $\pi' = v_0,\ldots, v_{n-1}$ and a word $u' \in \Delta^*$. Firstly, let's assume $v_{n-1} = \< q, q''>$ and $last(\pi) = \< q, q'>$. By definition, the edge $\< q, q''> \to \< q, q'>$ is possible only if an internal symbol or a return symbol is read, in order to keep the stack empty:

\begin{itemize}
\item if we read $a \in \Delta_i$ and have an empty stack, then the new $\weight_A$ is computed with the third case in equation~\eqref{eq:SWVPA-weight}.
We have
\begin{align*}
 \weight_A\bigl(\config{q}{\gamma}, u\, a,
               \config{q'}{\bot}\bigr) &=
   \weight_A\bigl(\config{q}{\gamma}, u, \config{q''}{\bot}\bigr)
   \otimes\weiei(q'', a, q')  \\
 &= \weight(\pi')\otimes\weiei(q'', a, q')\\
 &= \weight(\pi)\nonumber 
\end{align*}
with $\pi = (\pi', v_n)$, since we know that $\weight(v_0,\ldots, v_{n-1}) \otimes \eta_A(v_{n-1} \to v_{n}) = \weight(v_0,\ldots, v_n)$ and $\weiei(q'', a, q')$ is indeed a possible edge label $\eta_A$ for the edge $\< q, q''> \to \< q, q'>$ (first line of label definition). Hence in this case the Lemma is true.

\item Following the same reasoning as before, we find that the Lemma is also true when $a \in \Delta_r$ by using the last equation in \eqref{eq:SWVPA-weight} which illustrates the case when a return symbol is unmatched.
%
\end{itemize}
\noindent
Let's now assume that $v_{n-1} = \< q_1, p, q_2>$ and $last(\pi) = \< q_0, q_3>$. The edge $\< q_1, p, q_2> \to \< q_0, q_3>$ exists if we read a return then try to match with the call on top of the stack which then leaves the stack empty. With the 6th equation in~\eqref{eq:SWVPA-weight}, we can compute the $\weight_A$ for $\<q_0, q_3>$ and then replace the first term with the 5th equation in~\eqref{eq:SWVPA-weight}, such as:
\begin{align*}
&\weight_A\bigl(\config{q_0}{\bot}, u\,{\return{r}},
               \config{q_3}{\bot}\bigr) =
  \weight_A\bigl(\config{q_0}{\bot}, u,
                 \configup{q_2}{\<{\call{c}}, p>\stackup\bot}\bigr)
  \otimes \weir\bigl(q_2, {\call{c}}, p, {\return{r}}, q_3\bigr)\\
  &=
   \weight_A(\configup{q_1}{\<{\call{c}}, p>\stackup\bot}, u, \configup{q_2}{\<{\call{c}}, p>\stackup\bot}\bigr)
 \otimes \weiec(q_0, {\call{c}}, p, q_1)
  \otimes \weir\bigl(q_2, {\call{c}}, p, {\return{r}}, q_3\bigr)\\
  &= \weight_A(\pi')\otimes \weiec(q_0, {\call{c}}, p, q_1)
  \otimes \weir\bigl(q_2, {\call{c}}, p, {\return{r}}, q_3\bigr)\\
  &= \weight_A(\pi)
\end{align*}
for $\pi = (\pi', v_n)$. Here the Lemma holds since $\weir\bigl(q_2, {\call{c}}, p, {\return{r}}, q_3\bigr)$ is also part of the edge labels (second line).

Finally, let's assume that $v_{n-1} = \<q_1, p, q_2>$ and $last(\pi) = \<q_0, p', q_3>$ for $p, p' \in Q$. Following the same reasoning as the two previous cases, we can similarly prove that the Lemma is also true.
\qed
\end{proof}

@@

\noindent
For $n = n_{q, q'} > 0$, let $\bar\tau$ and~$\bar{w}$
be the sequences associated to $q$ and $q'$ as above.
Let us do a case analysis of $w_n$.
Computing $b_\bot(q, q')$ comes down to running $A$ on a well-matched word $u \in \Delta^*$ for which $|u| = n$. Let's take the case where $w_n  = \weiei$. Then using Equation~\ref{eq:SWVPA-weight}, we decompose $b_\bot(q, q')$ as follows:
%\config{q}{\gamma}, u\,{\call{c}'},
%\configup{q'}{\<{\call{c}'}, p'>\stackup \<{\call{c}}, p>\stackup \gamma'}\bigr)
\begin{align*}
b_\bot(q, q') &= \weight_A(\config{q}{\bot}, u', \config{q''}{\bot}) \otimes\weiei(q'', a, q') \\
&= b_\bot(q, q'')\otimes\weiei(q'', a, q')
\end{align*}
Since $u=u'a$ is minimal, then $u'$ is minimal, and $\weight_A(\config{q}{\bot}, u', \config{q''}{\bot}) =b_\bot(q, q'')$. Consequently, for this case the Lemma is true. The other cases can be proven similarly,  but only for the $w_n$ that can form a well-matched word. For example, $w_n$ cannot be $\weic$ since this call will never have a matching return.
\qed
\end{proof}

>>>>

% !TEX root = main.tex
Proposition~\ref{algo-shortest} states the correctness of the construction of the graph~$G_A$ for the computation of $b_\bot$
and, by extension, the minimum of $A$.
Its proof is a consequence of the two following Lemmata~\ref{algo-correct} and~\ref{algo-complete}.

@@

\noindent
If $\pi = v_0,\ldots, v_n$ with $n \geq 1$, 
we assume that Lemma~\ref{algo-correct} holds for a safe path $\pi' = v_0,\ldots, v_{n-1}$ and a word~$u' \in \Delta^*$, 
and do a case analysis on the edge~$v_{n-1} \to v_n$.

\smallskip\noindent
Firstly, let us consider the case where $v_{n-1} = \< q, q''>$ and $v_{n} = \< q, q'>$. 
By definition, the edge $\< q, q''> \to \< q, q'>$ is possible only if an internal symbol or a return symbol is read, 
keeping the stack empty:

\begin{itemize}
\item if we read $a \in \Delta_i$ and have an empty stack, 
then the new $\weight_A$ is computed 
with the third case in equation~\eqref{eq:SWVPA-weight}.
We have
\begin{align*}
 \weight_A\bigl(\config{q}{\gamma}, u'\, a,
               \config{q'}{\bot}\bigr) &=
   \weight_A\bigl(\config{q}{\gamma}, u', \config{q''}{\bot}\bigr)
   \otimes\weiei(q'', a, q')  \\
 &= \weight(\pi')\otimes\weiei(q'', a, q')\\
 &= \weight(\pi)\nonumber 
\end{align*}
since we know that $\weight(v_0,\ldots, v_{n-1}) \otimes \eta_A(v_{n-1} \to v_{n}) = \weight(v_0,\ldots, v_n)$ and $\weiei(q'', a, q')$ is indeed a possible label $\eta_A$ for the edge $\< q, q''> \to \< q, q'>$ (first line of label definition). 
%Hence in this case the Lemma is true.

\item Following the same reasoning as above, we find that the Lemma is also true when $a \in \Delta_r$ by using the last equation in \eqref{eq:SWVPA-weight} which corresponds to the case of an unmatched return symbol.
%
\end{itemize}
\noindent
let us now consider the case where $v_{n-1} = \< q_1, p, q_2>$ and $v_{n} = \< q_0, q_3>$. 
The edge $\< q_1, p, q_2> \to \< q_0, q_3>$ exists if we have a return symbol 
matching a call symbol on the top of the stack, 
which is pop, leaving the stack empty. 
With the 6th equation in~\eqref{eq:SWVPA-weight}, 
we can compute the $\weight_A$ for $\<q_0, q_3>$ 
and then replace the first term with the 5th equation in~\eqref{eq:SWVPA-weight}, such as:
\begin{align*}
&\weight_A\bigl(\config{q_0}{\bot}, u\,{\return{r}},
               \config{q_3}{\bot}\bigr) =
  \weight_A\bigl(\config{q_0}{\bot}, u,
                 \configup{q_2}{\<{\call{c}}, p>\stackup\bot}\bigr)
  \otimes \weir\bigl(q_2, {\call{c}}, p, {\return{r}}, q_3\bigr)\\
  &=
   \weight_A(\configup{q_1}{\<{\call{c}}, p>\stackup\bot}, u, \configup{q_2}{\<{\call{c}}, p>\stackup\bot}\bigr)
 \otimes \weiec(q_0, {\call{c}}, p, q_1)
  \otimes \weir\bigl(q_2, {\call{c}}, p, {\return{r}}, q_3\bigr)\\
  &= \weight_A(\pi')\otimes \weiec(q_0, {\call{c}}, p, q_1)
  \otimes \weir\bigl(q_2, {\call{c}}, p, {\return{r}}, q_3\bigr)\\
  &= \weight_A(\pi)
\end{align*}
%
Here the Lemma holds since $\weir\bigl(q_2, {\call{c}}, p, {\return{r}}, q_3\bigr)$ 
is also part of the edge labels (second line).

Finally, for the case where $v_{n-1} = \<q_1, p, q_2>$ and $v_n = \<q_0, p', q_3>$ for $p, p' \in Q$, 
we can following the same reasoning as the two previous cases, and complete the proof of Lemma~{algo-correct}.
\qed
\end{proof}

@@

\noindent
For $n = n_{q, q'} > 0$, let $\bar\tau$ and~$\bar{w}$
be the sequences associated to $q$ and $q'$ as above.
%
We can perform a case analysis of $w_n$.
Computing $b_\bot(q, q')$ comes down to running $A$ on a word $u \in \Delta^*$, 
such that $|u| = n$, and $u$ is well-matched except for unmatched return symbols
read on empty stack.
%
Let us consider the case where $w_n  = \weiei$. 
Then using Equation~\ref{eq:SWVPA-weight}, we can decompose $b_\bot(q, q')$ as follows:
%\config{q}{\gamma}, u\,{\call{c}'},
%\configup{q'}{\<{\call{c}'}, p'>\stackup \<{\call{c}}, p>\stackup \gamma'}\bigr)
\begin{align*}
b_\bot(q, q') &= \weight_A(\config{q}{\bot}, u', \config{q''}{\bot}) \otimes\weiei(q'', a, q') \\
&= b_\bot(q, q'')\otimes\weiei(q'', a, q')
\end{align*}
Since $u = u'a$ is minimal in length, then so is $u'$, 
and $\weight_A(\config{q}{\bot}, u', \config{q''}{\bot}) = b_\bot(q, q'')$. 
Consequently, for this case the Lemma holds. 
The other cases can be proven similarly,  but only for the $w_n$ that can form a well-matched word. 
For example, $w_n$ cannot be $\weic$ since this call would not have a matching return.
\qed
\end{proof}

### 2021-10-15 introduction.tex
submission FoSSaCS

% !TEX root = main.tex
%
% Introduction
%
Various extensions of language models have been proposed for handling infinite alphabets.
%words carrying data values in an infinite domain (e.g. integers) e.g. data processing
Some automata 
%\reviews{1) and grammar with storage?}
with memory extensions
allow restricted storage and comparison of input symbols
%and correspond with logics,
(see~\cite{Segoufin06csl} for a survey).
\florent{register: skip refs and details, add Mikolaj recent}
They use pebbles for marking positions~\cite{NevenSchwentickVianu04FSMinfinite},
registers~\cite{KaminskiFrancez94},
or %computations in 2 steps
separate computation on subsequences
with the same attribute values~\cite{Bojanczyk11FO2}. % data words automata.
%
%for the and verification of infinite state systems
%(model checkers: alphabet = long bit-vectors)
%...For the representation of  in model checking, verification and
%
Moreover, automata at the core of model checkers
compute on input symbols represented by large bitvectors~\cite{Vardi07ciaa} %\cite{BaierKatoen08MC}
%(sets of assignments of Boolean variables) %propositional variables)
and, in practice,  %implementation
each transition accepts a set of such symbols (instead of an individual symbol),
represented by Boolean Formulas or Binary Decision Diagrams.
%
Following this idea, % and generalizing,
in symbolic automata (\SA)~\cite{Veanes12symbolic,dAntoniVeanes17CAV,dAntoni21CACM},
transitions are guarded by predicates over large or infinite domains.
With appropriate closure conditions on the sets of such predicates, % (alphabet theories),
all the good properties enjoyed by automata over finite alphabets are preserved.
The ability to compare input symbols of \SA is quite restricted compared to the 
former automata with memory. 
It can however be extended with the addition of a stack~\cite{dAntonyAlur14SVPDA}.

Other extensions of language models  %(automata and grammars)
assign one weight value to every input~\cite{Droste09handbook}.
%help in dealing with non-determinism
This is useful for the quantitative modelling of
\eg probabilistic or stochastic recursive programs, %( for pushdown models),
quantitative database queries, or semi-structured data, % (in infinite domains), 
as well as for the verification of quantitative properties about 
\florent{quantitative reasoning/verification on programs}
quality measures, resource-consumption, distance metrics, probabilistic guarantees, \etc
%
In the context of parsing, when grammars return weight values, 
it is possible to rank derivations (and \emph{abstract syntax trees})
in order to select a best one (or $n$ bests),
%\florent{= "weighted parsing"}
\eg in case of ambiguity~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing}.
%This is roughly the principle of \emph{weighted parsing} approaches
%\reviews{1) jump from automata to grammars is uneasy and not really necessary}
% for which there may exist several derivations yielding one input word.
%
In \emph{weighted language models}
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA)~\cite{Droste09handbook},
a weight is associated to each transition rule, % production rule
and the weights of the rules involved in a computation are combined with an
associative product operator~$\otimes$. % to yield the weight of a whole computation.
%\reviews{1) AST for grammar, run for WA}
A second operator~$\oplus$
is moreover used to resolve the ambiguity raised by the existence
of several (in general exponentially many) computations on a given input.
Typically, $\oplus$ selects the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators is typically 
a \emph{semiring} %$\Semiring$
where $\oplus$ can be extended to infinite sums,
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra% % see Figure~\ref{fig:semirings}.
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$,
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

In this paper, we present some results for 
\emph{Symbolic Weighted} finite states language models %(\swM) 
generalizing the Boolean guards of~\SA %in transitions
to functions into an arbitrary semiring, 
and also the~\WA, by handling infinite alphabets (Figure~\ref{fig:hierarchy}).
%
The models considered are also 
%\SWA and \SWVPA are both 
particular cases of the very general class of 
Weighted Symbolic Automata with Data Storage~\cite{Herrmann16dlt,Herrmann20phd}, 
using appropriate storage types.
%with special data storage types...
%\florent{Both WA and SA generalized by \cite{Herrmann16dlt,Herrmann20phd}...}
%\reviews{cite \cite{Herrmann16dlt,Herrmann20phd}}

@@

\noindent
In short, a transition rule $q \xrightarrow{\phi} q'$ of a SW automaton, 
from state $q$ to $q'$,
%\florent{c'est un résumé de la Fig. \ref{fig:hierarchy}}
%input symbols are variables, %(parameters)
is labeled by a function~$\phi$ associating to every input symbol~$a$, a weight value~$\phi(a)$
in a semiring~$\Semiring$.
%
The models studied are: 
symbolic-weighted automata (\SWA),
defining series over infinite alphabets, 
transducers (\SWT), 
defining distances between finite words over infinite alphabets, % following~\cite{Mohri03ijfcs}.
and pushdown automata with a visibility restriction~\cite{AlurMadhusudan09nested} (\SWVPA),
operating sequentially on %\emph{nested words}~\cite{AlurMadhusudan09nested},
words structured with markup symbols (parentheses) and describing linearizations of trees.
%A \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$
%to a given nested word $t$, which is itsef the linearization of a weighted AST. %representing a parse tree.
%
The main contributions of this paper are:
%\reviews{1) not $i$! see response.txt}
%($i$)~the introduction of automata, \SWA, transducers, \SWT (Section~\ref{section:SWA}),
%and visibly pushdown automata \SWVPA (Section~\ref{sec:SWVPA}),
%generalizing the corresponding classes of symbolic and weighted models,
%for the weighted computation on (nested) words over infinite alphabets;
($i$)~the construction à la Bar-Hillel Perles Shamir of a \SWA
     computing a \SWT-defined distance between a \SWA input language and a word (Section~\ref{section:SWA}), 
($ii$)~closure results for \SWVPA, 
($iii$)~a polynomial best-search algorithm for \SWVPA (Section~\ref{sec:SWVPA}). %(Section~\ref{sec:best})
% a framework for parsing over infinite alphabets,
Moreover, we present in Section~\ref{sec:parsing} an 
($iv$) application to the problem of weighted parsing over infinite input alphabets, 
called \emph{SW-parsing}. 
%
The goal of this problem is, given an input word~$s$, 
to find~$t$ minimizing the distance, in the sense of~\cite{Mohri03ijfcs}, 
$T(s, t) \otimes A(t)$, where $T$ is a \SWT and $A$ a \SWVPA.
The notion of transducer-based distances allows to consider 
different infinite alphabets for the input $s$ and output $t$.
Moreover, the use of \SWVPA permits to consider output $t$ in the form of a (nested) word, instead of a tree.
%technical convenience
%the SW models are integrated in a uniform framework.
\emph{SW-parsing} is solved with the Bar-Hillel construction~\cite{NederhofSatta03ParsingIntersection} 
%\cite{GruneJacobs08parsing}
of a \SWVPA $B$ such that, for all~$t$, $B(t) = T(s, t) \otimes A(t)$
and the application of a best-search procedure to this automaton $B$.
% reduction to the shortest distance in graphs~\cite{Mohri02semiring,Huang05kbest}.

@@

%Parsing is the problem of structuring a linear representation
%(a finite word) according to a language model. % (a formal grammar). % natural language, programming language,
Context-free parsing approaches~\cite{GruneJacobs08parsing}
generally assume a finite and reasonably small input alphabet. %models and algorithms
%Such a restriction makes perfect sense in the context of
%NLP tasks such as constituency parsing, 
%compilers or interpreters for programming languages.
Considering large or infinite alphabets can however be of
practical interest when dealing with large characters encodings such as UTF-16~\cite{dAntoni21CACM}.
\florent{applis à la Fossacs chez Veanes et al?}
% processing strings in \eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM},
%We believe that these results could also be useful 
It is also true in the context of automata-based quantitative verification,  %of systems processing %analysis of 
\eg when dealing with data streams, 
serialization of structured documents~\cite{Segoufin06csl,NevenSchwentickVianu04FSMinfinite},
or timed execution traces~\cite{Bouyer03algebraic}.

@@

The latter case is related to a motivation study for the present work:
automated music transcription, 
that we shall use as a running example
thorough the paper, in order to illustrate our results.
%
Representations capturing music performances
are essentially linear~\cite{Selfridge-Field97beyondMIDI};
%audio files or the widely used MIDI format %\cite{MIDIfile}.
they ignore the hierarchical structures that frame the
conception of music, at least in the Western area. 
These structures, on the other hand, are present, either explicitly  or implicitly,
in Common Western Music Notation~\cite{Gould11Notation}:
%in music notation~\cite{Gould11Notation}:
Music scores are partitioned in measures,
measures in beats, and beats can be further recursively divided.
It follows that written music events do not occur at arbitrary timestamps,
but respect a discrete partitioning of the timeline incurred by
these recursive divisions.
The \emph{transcription problem} takes
as input a linear performance (in audio or MIDI format) 
and aims at re-constructing structured notation, 
by mapping input events to this hierarchical rhythmic space.
It can therefore be stated as a parsing problem
over an infinite alphabet of timed events~\cite{foscarin:hal-01988990}.

>>>>

% !TEX root = main.tex
%
% Introduction
%
Various extensions of language models have been proposed for handling infinite alphabets.
%words carrying data values in an infinite domain (e.g. integers) e.g. data processing
Some automata 
%\reviews{1) and grammar with storage?}
with memory extensions
allow restricted storage and comparison of input symbols
%and correspond with logics,
(see~\cite{Segoufin06csl} for a survey).
\florent{register: skip refs and details, add Mikolaj recent}
They use pebbles for marking positions~\cite{NevenSchwentickVianu04FSMinfinite},
registers~\cite{KaminskiFrancez94},
or %computations in 2 steps
separate computation on subsequences
with the same attribute values~\cite{Bojanczyk11FO2}. % data words automata.
%
%for the and verification of infinite state systems
%(model checkers: alphabet = long bit-vectors)
%...For the representation of  in model checking, verification and
%
Moreover, automata at the core of model checkers
compute on input symbols represented by large bitvectors~\cite{Vardi07ciaa} %\cite{BaierKatoen08MC}
%(sets of assignments of Boolean variables) %propositional variables)
and, in practice,  %implementation
each transition accepts a set of such symbols (instead of an individual symbol),
represented by Boolean Formulas or Binary Decision Diagrams.
%
Following this idea, % and generalizing,
in symbolic automata (\SA)~\cite{Veanes12symbolic,dAntoniVeanes17CAV,dAntoni21CACM},
transitions are guarded by predicates over large or infinite domains.
With appropriate closure conditions on the sets of such predicates, % (alphabet theories),
all the good properties enjoyed by automata over finite alphabets are preserved.
The ability of \SA to compare input symbols is restricted compared to the 
former automata with memory. 
It can however be extended with the addition of a stack~\cite{dAntonyAlur14SVPDA}.

Other extensions of language models  %(automata and grammars)
assign one weight value to every input~\cite{Droste09handbook}.
%help in dealing with non-determinism
This is useful for the quantitative modelling of
\eg probabilistic or stochastic recursive programs, %( for pushdown models),
quantitative database queries, or semi-structured data, % (in infinite domains), 
as well as for the verification of quantitative properties of systems
related to %\florent{quantitative reasoning/verification on programs}
quality measures, resource-consumption, distance metrics, probabilistic guarantees, \etc
%
In the context of parsing, when grammars return weight values, 
it is possible to rank derivations (and \emph{abstract syntax trees})
in order to select a best one (or $n$ bests),
%\florent{= "weighted parsing"}
\eg in case of ambiguity~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing}.
%This is roughly the principle of \emph{weighted parsing} approaches
%\reviews{1) jump from automata to grammars is uneasy and not really necessary}
% for which there may exist several derivations yielding one input word.
%
In \emph{weighted language models}
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA)~\cite{Droste09handbook},
a weight is associated to each transition rule, % production rule
and the weights of the rules involved in a computation are combined with an
associative product operator~$\otimes$. % to yield the weight of a whole computation.
%\reviews{1) AST for grammar, run for WA}
A second operator~$\oplus$
is moreover used to resolve the ambiguity raised by the existence
of several (in general exponentially many) computations on a given input.
Typically, $\oplus$ selects the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators is typically 
a \emph{semiring} %$\Semiring$
where $\oplus$ can be extended to infinite sums,
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra% % see Figure~\ref{fig:semirings}.
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$,
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

In this paper, we present some results for 
\emph{Symbolic Weighted} finite states language models %(\swM) 
generalizing the Boolean guards of~\SA %in transitions
to functions into an arbitrary semiring, 
and also the~\WA, by handling infinite alphabets (Figure~\ref{fig:hierarchy}).
%
In short, a transition rule $q \xrightarrow{\phi} q'$ of a Symbolic Weighted Automaton, 
from state $q$ to $q'$,
%\florent{c'est un résumé de la Fig. \ref{fig:hierarchy}}
%input symbols are variables, %(parameters)
is labeled by a function~$\phi$ associating to every input symbol~$a$, a weight value~$\phi(a)$
in a semiring~$\Semiring$.
%
These models are also 
%\SWA and \SWVPA are both 
particular cases of the very general class of 
Weighted Symbolic Automata with Data Storage~\cite{Herrmann16dlt,Herrmann20phd}, 
for appropriate storage types.
%with special data storage types...
%\florent{Both WA and SA generalized by \cite{Herrmann16dlt,Herrmann20phd}...}
%\reviews{cite \cite{Herrmann16dlt,Herrmann20phd}}

@@

\noindent
The models studied are: 
symbolic-weighted automata (\SWA),
defining series over infinite alphabets, 
transducers (\SWT), 
defining distances between finite words over infinite alphabets, % following~\cite{Mohri03ijfcs}.
and pushdown automata with a visibility restriction~\cite{AlurMadhusudan09nested} (\SWVPA),
operating sequentially on %\emph{nested words}~\cite{AlurMadhusudan09nested},
words structured with markup symbols (parentheses) and describing linearizations of trees.
%A \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$
%to a given nested word $t$, which is itsef the linearization of a weighted AST. %representing a parse tree.
%
The main contributions of this paper are:
%\reviews{1) not $i$! see response.txt}
%($i$)~the introduction of automata, \SWA, transducers, \SWT (Section~\ref{section:SWA}),
%and visibly pushdown automata \SWVPA (Section~\ref{sec:SWVPA}),
%generalizing the corresponding classes of symbolic and weighted models,
%for the weighted computation on (nested) words over infinite alphabets;
\begin{description}
\item[($i$)] a construction à la Bar-Hillel Perles Shamir of a \SWA
     computing a \SWT-defined distance between a \SWA input language and a word (Proposition~\ref{prop:epsilon}), 
\item[($ii$)] closure results for \SWVPA (Proposition~\ref{prop:SWVPA-product}), 
\item[($iii$)] a polynomial best-search algorithm for \SWVPA (Proposition~{th:best-search}. %(Section~\ref{sec:best})
\end{description}
%
% a framework for parsing over infinite alphabets,
Moreover, we present in Section~\ref{sec:parsing} an 
\begin{description}
\item[($iv$)] application to the problem of weighted parsing over infinite input alphabets, 
called \emph{SW-parsing}. 
\end{description}
%
The goal of the latter problem is, given an input word~$s$, 
to find~$t$ minimizing the distance, in the sense of~\cite{Mohri03ijfcs}, 
$T(s, t) \otimes A(t)$, where $T$ is a \SWT and $A$ a \SWVPA.
The notion of transducer-based distances allows to consider 
different infinite alphabets for the input $s$ and output $t$.
Moreover, the use of \SWVPA permits to search for an output $t$ in the form of a (nested) word, 
instead of a tree.
%technical convenience
%the SW models are integrated in a uniform framework.
\emph{SW-parsing} is solved with a Bar-Hillel construction~\cite{NederhofSatta03ParsingIntersection} 
%\cite{GruneJacobs08parsing}
of a \SWVPA $B$ such that, for all~$t$, $B(t) = T(s, t) \otimes A(t)$
and the application of a best-search procedure to this automaton $B$.
% reduction to the shortest distance in graphs~\cite{Mohri02semiring,Huang05kbest}.

@@

%Parsing is the problem of structuring a linear representation
%(a finite word) according to a language model. % (a formal grammar). % natural language, programming language,
Context-free parsing approaches~\cite{GruneJacobs08parsing}
generally assume a finite and reasonably small input alphabet. %models and algorithms
%Such a restriction makes perfect sense in the context of
%NLP tasks such as constituency parsing, 
%compilers or interpreters for programming languages.
Considering large or infinite alphabets can however be of
practical interest when dealing with large characters encodings such as UTF-16~\cite{dAntoni21CACM}.
% processing strings in \eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM},
%We believe that these results could also be useful 
It is also true in the context of automata-based quantitative verification techniques~\cite{Kwiatkowska07quantitativeVerif},  %of systems processing %analysis of 
in particular when dealing with data streams, 
serialization of structured documents~\cite{Segoufin06csl,NevenSchwentickVianu04FSMinfinite},
or timed execution traces~\cite{Bouyer03algebraic}.

@@

The latter case of timed traces 
is related to a problem that motivated the present work:
automated music transcription, 
that we shall use as a running example
thorough the paper, in order to illustrate our results.
%
Representations capturing music performances
are essentially linear~\cite{Selfridge-Field97beyondMIDI};
%audio files or the widely used MIDI format %\cite{MIDIfile}.
they ignore the hierarchical structures that frame the
conception of music, at least in the Western area. 
Hierarchical structures, on the other hand, are present, either explicitly  or implicitly,
in Common Western Music Notation~\cite{Gould11Notation}:
%in music notation~\cite{Gould11Notation}:
Music scores are partitioned in measures,
measures in beats, and beats can be further recursively divided.
It follows that written music events do not occur at arbitrary timestamps,
but respect a discrete partitioning of the timeline incurred by
these recursive divisions.
The \emph{transcription problem} takes
as input a linear performance (in audio or MIDI format) 
and aims at re-constructing structured notation, 
by mapping input events to this hierarchical rhythmic space.
It can therefore be stated as a parsing problem
over an infinite alphabet of timed events~\cite{foscarin:hal-01988990}.
